---
title: "Imputing missing timepoints in longitudinal microbiome data"
author:
- name: Alireza Dostmohammadi
  email: alireza.dostmohammadi@uni-due.de
package: TaxaTimeImpute
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Imputing missing microbiome timepoints}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4
)
```

# Introduction

Longitudinal microbiome studies often lose whole samples. A subject misses a
visit, or a library fails quality control, and the affected subject-timepoint
is absent from the abundance table. Removing those subjects discards usable
data, and leaving the gap affects any downstream calculation.

`r Biocpkg("TaxaTimeImpute")` estimates the missing values. It models each
taxon's abundance over time as a smooth curve and shares information across
subjects using Functional Principal Component Analysis (FPCA) in the PACE
formulation. The covariance surface is estimated from all subjects together,
so sparse and irregularly sampled designs are supported. Per-subject smoothing
is not possible for such designs.

The package handles missing whole samples. A sample column must therefore be
either fully observed or entirely `NA`; a column holding values for some taxa
and `NA` for others is a different problem, and is reported as an error rather
than imputed.

```{r load}
library(TaxaTimeImpute)
data(taxa_demo)
```

# Giving the package your own tables

The usual starting point is the pair of tables a study already produces: an
abundance table with taxa in rows and one column per sample, and a metadata
sheet saying which sample came from which subject at which time. You say
which metadata column is which, so they can be named anything.

```{r own-tables}
counts <- data.frame(
    taxon = c("Bacteroides", "Prevotella", "Akkermansia"),
    M01_d0 = c(1.20, 0.41, -0.3), M01_d7 = c(1.51, 0.62, -0.1),
    M02_d0 = c(0.98, 0.55, 0.2), M02_d7 = c(1.14, 0.47, 0.4),
    M03_d0 = c(1.05, 0.51, 0.1)
)

meta <- data.frame(
    sample = c("M01_d0", "M01_d7", "M02_d0", "M02_d7", "M03_d0"),
    mouse = c("M01", "M01", "M02", "M02", "M03"),
    day = c(0, 7, 0, 7, 0)
)
```

`tti_from_metadata()` converts the two into the form the model needs and
reports what the design is missing. Here mouse M03 has no day-7 sample:

```{r own-design}
design <- tti_from_metadata(
    counts, meta,
    sample_col = "sample", subject_col = "mouse", time_col = "day",
    taxon_col = "taxon"
)

design$missing
```

Everything printed is also returned, so nothing is available only by reading
the console.

The word *subject* is used throughout for whatever the repeated measurements
were taken on: a mouse, a participant, a plot, a bioreactor. Nothing in the
package assumes a species.

Passing the same arguments to `tti_run()` does the whole job and returns the
completed table under your own sample names. That is the recommended route,
and the rest of this vignette uses the older layout only because the bundled
example data is stored that way.

# The internal layout

Internally, and in the bundled example, a table has one column of taxon
identifiers and one column per sample named `<subject>.<time>`:

```{r peek}
taxa_demo[seq_len(4), seq_len(5)]
```

`taxa_demo` is simulated, with 12 taxa, 10 subjects and 7 time points. It
carries both forms of missingness the package recognises. A table in this
form can be passed directly, without metadata.

# What is missing, and why

Before fitting anything, ask what the table is actually missing:

```{r detect}
info <- tti_detect_missing(taxa_demo, taxon_col = "OTU_ID")
info$missing
```

Two reasons appear:

* `absent_column`: the subject-timepoint has no column.
* `all_na`: the column exists but every taxon in it is `NA`.

A sample column must be either fully observed or entirely `NA`. A column
holding values for some taxa and `NA` for others is neither, and is rejected
with an error naming the offending columns, so `partial_na` is empty in any
result you actually receive:

```{r partial}
info$partial_na
```

If that error appears, correct the data first: make the column entirely `NA`
if the sample is genuinely missing, or fill in the individual `NA` values if
it is not.

It is also worth checking how many observations each subject retains. A
subject with fewer than two informative points contributes to the model fit
but has little of its own trajectory to condition on:

```{r observed}
head(info$observed)
```

# Imputing a table

`tti_run()` does the whole job: it detects what is missing, fits the model and
returns the table with the gaps filled.

```{r run}
run <- tti_run(taxa_demo, taxon_col = "OTU_ID", K = 1, verbose = FALSE)
run
```

The completed table keeps the input layout. Existing columns keep their order
and their values, because imputation does not modify measured data:

```{r completed}
run$completed[seq_len(4), seq_len(5)]
```

Every imputed cell is also available in long form:

```{r long}
head(run$imputed[, c("species", "rep", "time", "imputed_value")])
```

## A note on warnings

Fitting sparse designs makes `r CRANpkg("fdapace")` emit
`There is a time gap of at least ...`. This describes the sampling design and
is not an error.

# Working with SummarizedExperiment

Most Bioconductor microbiome workflows carry data in a
`r Biocpkg("TreeSummarizedExperiment")`. `tti_run()` has a method for that,
reading subject and time from `colData`:

```{r se, message = FALSE}
library(SummarizedExperiment)

se <- tti_as_demo_se()
se
```

```{r se-run}
out <- tti_run(
    se,
    subject_col = "subject",
    time_col = "timepoint",
    K = 1,
    verbose = FALSE
)

assayNames(out)
```

The input assay is never modified. The completed matrix arrives as a new
assay, so observed and imputed values stay distinguishable:

```{r se-assays}
assay(out, "imputed")[seq_len(3), seq_len(4)]
```

What happened is recorded in the object's metadata:

```{r se-meta}
metadata(out)$tti_run$missing
metadata(out)$tti_run$added_samples
```

If a subject-timepoint had no sample at all, there is no column to write the
imputed values into, so one is appended and `added_samples` names it. The
returned object can therefore be wider than the input.

Subject identifiers and time points are recoded internally, so names may
contain dots or spaces and time points need not be integers or evenly spaced.

# Benchmarking: masking values you already have

To measure accuracy you need ground truth, so you hide values you already
have and score the predictions. That is a different entry point:
`tti_prepare()` takes an explicit mask.

```{r bench}
prep <- tti_prepare(
    dat = taxa_demo,
    taxon_col = "OTU_ID",
    mask_list = data.frame(rep = "S01", time = 3)
)

fit <- tti_fit(prep, K = 1)
imputed <- tti_impute(fit)

head(imputed[, c("species", "true_value", "imputed_value")])
```

Now `true_value` is populated, so error metrics mean something:

```{r metrics}
metrics <- tti_metrics(fit)
c(MSE = metrics$MSE_overall, RMSE = metrics$RMSE_overall)
```

After `tti_run()` the true values were never observed, so `true_value` and
`se` are `NA` and `tti_metrics()` is not meaningful. The two entry points
answer different questions.

# Uncertainty

`tti_ci()` propagates the FPCA score covariance to an analytic interval:

```{r ci}
ci <- tti_ci(
    fit = fit,
    species_name = imputed$species[1],
    rep_id = "S01",
    time_id = 3,
    method = "analytic"
)

head(ci$analytic)
```

A bootstrap interval is also available via `method = "bootstrap"`, which
refits the model `B` times and is correspondingly slower.

`tti_plot()` draws the trajectory with its band, the observed points, and the
imputed value:

```{r alt}
alt_text <- paste(
    "FPCA trajectory for one taxon with its analytic confidence band,",
    "the observed points, and the imputed value marked."
)
```

```{r plot, fig.alt = alt_text}
tti_plot(
    fit = fit,
    species_name = imputed$species[1],
    rep_id = "S01",
    time_id = 3,
    ci_method = "analytic"
)
```

Two panels are returned: the fitted curve with its band on the left, and the
subject's raw observations on the right.

Reading it:

* The **imputed value** sits on the fitted curve at the marked time. It is not
  interpolated between neighbouring points; it is where the shared
  time-course, conditioned on this subject's own observations, places the
  subject.
* The **width of the band** is the honest part. It widens where the subject
  has few nearby observations, so a narrow band means the estimate is well
  determined and a wide one means it rests mostly on the population curve.
  Compare it against the spread of the observed points: a band much wider than
  that spread is a weakly informed imputation, and worth reporting as such.
* **Observed points lying far from the curve** mean this subject departs from
  the shape shared across subjects. The imputation still borrows that shape,
  so treat such a value with more caution than the band alone suggests.

# Optional refinements

Two refinements are available, both switched through `tti_fit()`:

* **Functional outlier screening** (`use_outliers = TRUE`, the default) drops
  trajectories below the 5% Fraiman-Muniz depth quantile before refitting. The
  target subject is always kept.
* **Trajectory clustering** (`K = NULL`) selects the number of clusters per
  taxon by mean silhouette width and refits within the target's cluster.
  `K = 1` pools all subjects and is considerably faster.

Both were benchmarked in the accompanying publication. Outlier screening
improved 20 of 121 masks on one cohort and reduced accuracy above 0.3 masking.
Clustering did not improve accuracy on the masks tested. They are provided as
diagnostic options, not as default improvements.

# Reproducibility

`tti_fit()` takes a `seed`, and clustering is reproducible for a given seed.
The caller's own random stream is restored when the function returns, so
calling it does not disturb any other random draws in your session.

# Session information

```{r session}
sessionInfo()
```
