<!-- Shared body of the workflow vignette.

     Kept outside vignettes/ so R does not treat it as a vignette in its
     own right. Both the HTML and the PDF vignette include it as a knitr
     child, so the two formats cannot drift apart. -->

# Introduction

Longitudinal microbiome studies often lose whole samples. A subject misses a
visit, or a library fails quality control, and the affected subject-timepoint
is absent from the abundance table. Removing those subjects discards usable
data, and leaving the gap affects any downstream calculation.

`r Biocpkg("MicrobiomeCurves")` estimates the missing values. It models each
taxon's abundance over time as a smooth curve and shares information across
subjects using Functional Principal Component Analysis (FPCA) in the PACE
formulation. The covariance surface is estimated from all subjects together,
so sparse and irregularly sampled designs are supported. Per-subject smoothing
is not possible for such designs.

The package handles missing whole samples. A sample column must therefore be
either fully observed or entirely `NA`; a column holding values for some taxa
and `NA` for others is a different problem, and is reported as an error rather
than imputed.

```{r load}
library(MicrobiomeCurves)
data(taxa_demo)
```

# Giving the package your own tables

The input is the pair of tables a study already produces: an abundance table
with **taxa in the row names** and one column per sample, and a metadata
sheet saying which sample came from which subject at which time. You say
which metadata column is which, so they can be named anything, and nothing is
encoded in a column name.

The bundled example is available in exactly that form:

```{r own-tables}
demo <- mc_demo_data()

demo$counts[1:3, 1:4]
head(demo$metadata, 4)
```

`mc_from_metadata()` converts the two into the form the model needs and
reports what the design is missing:

```{r own-design}
design <- mc_from_metadata(
    demo$counts, demo$metadata,
    sample_col = "sample", subject_col = "subject", time_col = "time"
)

design$missing
```

Everything printed is also returned, so nothing is available only by reading
the console.

The word *subject* is used throughout for whatever the repeated measurements
were taken on: a mouse, a participant, a plot, a bioreactor. Nothing in the
package assumes a species.

## Time points

A study that records visits as labels can pass those instead of numbers:

```{r labels}
lab <- demo$metadata
lab$time <- factor(
    paste0("visit", lab$time),
    levels = paste0("visit", sort(unique(lab$time)))
)

d2 <- suppressWarnings(mc_from_metadata(
    demo$counts, lab,
    sample_col = "sample", subject_col = "subject", time_col = "time",
    verbose = FALSE
))

d2$axis$levels
d2$axis$positions
```

Labels are placed at consecutive positions, in the order given: the factor's
levels if it is a factor, the order the rows appear otherwise. A warning
states the order so it can be checked.

It is worth being clear about what the model does with time. **It uses the
order of the time points, not their values.** Days 0, 7 and 14 are fitted
exactly as 0, 1 and 2 would be, so a study with uneven intervals is not
represented as uneven in the fitted curve. Passing numbers rather than
labels does not change that; the values are kept for reporting, for the axis
of the uncertainty pages, and for naming any column that has to be created.

## Raw counts or CLR

The model works on centred log-ratios. If the table holds raw counts or
relative abundances, say so and the transformation is done here, replacing
zeros first:

```{r raw, eval = FALSE}
run <- mc_run(
    counts, meta,
    sample_col = "library", subject_col = "animal", time_col = "day",
    abundance_type = "raw"
)
```

## Writing results to disk

Give `out_dir` and the run writes three things:

| file | contents |
| --- | --- |
| `imputed_abundance.tsv` | the completed table |
| `imputation_log.txt` | design, time order, gaps, warnings |
| `uncertainty_by_taxon.pdf` | one page per taxon, with intervals |

```{r outdir}
out <- file.path(tempdir(), "mc_example")
run <- suppressWarnings(mc_run(
    demo$counts, demo$metadata,
    sample_col = "sample", subject_col = "subject", time_col = "time",
    out_dir = out, K = 1, verbose = FALSE
))

basename(run$files)
cat(head(readLines(file.path(out, "imputation_log.txt")), 12), sep = "
")
```

The uncertainty pages are what to look at before treating imputed values as
data. Each page shows every value imputed for one taxon with its interval, so
a taxon whose gaps rest on very little information declares itself:

```{r unc}
mc_uncertainty(run, rownames(demo$counts)[1])
```

`plots = FALSE` turns the drawing off. `plot_format = "png"` (or `"both"`)
writes one image per taxon at `dpi`, which defaults to 300; the PDF is vector
art and stays sharp at any size, so `dpi` does not apply to it.

Results come back under your own sample names, ordered by subject and then
time:

```{r completed}
run$completed[1:3, 1:5]
```

A column is added only for a subject-timepoint that had no sample. It is
named from its subject and time, and flagged in the returned metadata:

```{r added}
run$metadata[run$metadata$imputed, ]
```

# The internal layout

The fitting functions `mc_prepare()`, `mc_fit()` and `mc_impute()` work on
a wide table whose sample columns are named `<subject>.<time>`. That is the
benchmarking route, described later, and the form `taxa_demo` is stored in:

```{r peek}
taxa_demo[seq_len(4), seq_len(5)]
```

`taxa_demo` is simulated, with 12 taxa, 10 subjects and 7 time points. It
carries both forms of missingness the package recognises.

# What is missing, and why

Before fitting anything, ask what the table is actually missing:

```{r detect}
info <- mc_detect_missing(taxa_demo, taxon_col = "OTU_ID")
info$missing
```

Two reasons appear:

* `absent_column`: the subject-timepoint has no column.
* `all_na`: the column exists but every taxon in it is `NA`.

A sample column must be either fully observed or entirely `NA`. A column
holding values for some taxa and `NA` for others is neither, and is rejected
with an error naming the offending columns, so `partial_na` is empty in any
result you actually receive:

```{r partial}
info$partial_na
```

If that error appears, correct the data first: make the column entirely `NA`
if the sample is genuinely missing, or fill in the individual `NA` values if
it is not.

It is also worth checking how many observations each subject retains. A
subject with fewer than two informative points contributes to the model fit
but has little of its own trajectory to condition on:

```{r observed}
head(info$observed)
```

# Imputing a table

`mc_run()` does the whole job: it converts the tables, reports what is
missing, fits the model, and returns the table with the gaps filled. It was
run above; the pieces of its result are:

```{r run}
run
```

Every imputed cell is also available in long form, under your own subject and
sample names rather than anything internal:

```{r long}
head(run$imputed[
    , c("species", "subject", "sample", "time", "imputed_value")
])
```

## A note on warnings

Fitting sparse designs makes `r CRANpkg("fdapace")` emit
`There is a time gap of at least ...`. This describes the sampling design and
is not an error.

# Working with SummarizedExperiment

Most Bioconductor microbiome workflows carry data in a
`r Biocpkg("TreeSummarizedExperiment")`. `mc_run()` has a method for that,
reading subject and time from `colData`:

```{r se, message = FALSE}
library(SummarizedExperiment)

se <- mc_as_demo_se()
se
```

```{r se-run}
out <- mc_run(
    se,
    subject_col = "subject",
    time_col = "timepoint",
    K = 1,
    verbose = FALSE
)

assayNames(out)
```

The input assay is never modified. The completed matrix arrives as a new
assay, so observed and imputed values stay distinguishable:

```{r se-assays}
assay(out, "imputed")[seq_len(3), seq_len(4)]
```

What happened is recorded in the object's metadata:

```{r se-meta}
metadata(out)$mc_run$missing
metadata(out)$mc_run$added_samples
```

If a subject-timepoint had no sample at all, there is no column to write the
imputed values into, so one is appended and `added_samples` names it. The
returned object can therefore be wider than the input.

Subject identifiers and time points are recoded internally, so names may
contain dots or spaces and time points need not be integers or evenly spaced.

# Benchmarking: masking values you already have

To measure accuracy you need ground truth, so you hide values you already
have and score the predictions. That is a different entry point:
`mc_prepare()` takes an explicit mask.

```{r bench}
prep <- mc_prepare(
    dat = taxa_demo,
    taxon_col = "OTU_ID",
    mask_list = data.frame(rep = "S01", time = 3)
)

fit <- mc_fit(prep, K = 1)
imputed <- mc_impute(fit)

head(imputed[, c("species", "true_value", "imputed_value")])
```

Now `true_value` is populated, so error metrics mean something:

```{r metrics}
metrics <- mc_metrics(fit)
c(MSE = metrics$MSE_overall, RMSE = metrics$RMSE_overall)
```

After `mc_run()` the true values were never observed, so `true_value` and
`se` are `NA` and `mc_metrics()` is not meaningful. The two entry points
answer different questions.

# Uncertainty

`mc_ci()` propagates the FPCA score covariance to an analytic interval:

```{r ci}
ci <- mc_ci(
    fit = fit,
    species_name = imputed$species[1],
    rep_id = "S01",
    time_id = 3,
    method = "analytic"
)

head(ci$analytic)
```

A bootstrap interval is also available via `method = "bootstrap"`, which
refits the model `B` times and is correspondingly slower.

`mc_plot()` draws the trajectory with its band, the observed points, and the
imputed value:

```{r alt}
alt_text <- paste(
    "FPCA trajectory for one taxon with its analytic confidence band,",
    "the observed points, and the imputed value marked."
)
```

```{r plot, fig.alt = alt_text}
mc_plot(
    fit = fit,
    species_name = imputed$species[1],
    rep_id = "S01",
    time_id = 3,
    ci_method = "analytic"
)
```

Two panels are returned: the fitted curve with its band on the left, and the
subject's raw observations on the right.

Reading it:

* The **imputed value** sits on the fitted curve at the marked time. It is not
  interpolated between neighbouring points; it is where the shared
  time-course, conditioned on this subject's own observations, places the
  subject.
* The **width of the band** is the honest part. It widens where the subject
  has few nearby observations, so a narrow band means the estimate is well
  determined and a wide one means it rests mostly on the population curve.
  Compare it against the spread of the observed points: a band much wider than
  that spread is a weakly informed imputation, and worth reporting as such.
* **Observed points lying far from the curve** mean this subject departs from
  the shape shared across subjects. The imputation still borrows that shape,
  so treat such a value with more caution than the band alone suggests.

# Optional refinements

Two refinements are available, both switched through `mc_fit()`:

* **Functional outlier screening** (`use_outliers = TRUE`, the default) drops
  trajectories below the 5% Fraiman-Muniz depth quantile before refitting. The
  target subject is always kept.
* **Trajectory clustering** (`K = NULL`) selects the number of clusters per
  taxon by mean silhouette width and refits within the target's cluster.
  `K = 1` pools all subjects and is considerably faster.

Both were benchmarked in the accompanying publication. Outlier screening
improved 20 of 121 masks on one cohort and reduced accuracy above 0.3 masking.
Clustering did not improve accuracy on the masks tested. They are provided as
diagnostic options, not as default improvements.

# Reproducibility

`mc_fit()` takes a `seed`, and clustering is reproducible for a given seed.
The caller's own random stream is restored when the function returns, so
calling it does not disturb any other random draws in your session.

# Session information

```{r session}
sessionInfo()
```
